Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-10 paper/paper.tex
Stuff

\subsection{Types and Morphisms}
\label{sec:types-morphisms}
We address the different forms of equality by defining a class of
\emph{setoids}, types with an associated equivalence relation, and considering
only functions that respect the associated equivalences, or \emph{proper}
morphisms with respect to the function \emph{respectfulness} relation. We use
the type-class mechanism, instead of setoids in Coq's standard library, to help
Coq's code extraction mechanism remove any occurence of custom equivalence
relations in the extracted OCaml code. We use \coq{=e} to denote the equivalence
relation of a setoid. By default, we associate every Coq type with the standard
propositional equality, unless a different equivalence is specified (we allow
overlapping instances, and Coq's propositional equality takes the lowest
priority).
%\begin{coqcode}
%Reserved Notation "f =e g" (at level 70, no associativity).
%Class setoid A : Type :=
%  MkEquiv
%    { eqRel : A -> A -> Prop;
%      e_refl : forall x, eqRel x x;
%      e_sym : forall x y, eqRel x y -> eqRel y x;
%      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
%    }.
%Notation "f =e g" := (eqRel f g).
%\end{coqcode}
Given types \coq{A} and  \coq{B}, with their respective
equivalence relations \coq{eA : A -> A -> Prop} and
\coq{eB : B -> B -> Prop}, the we define the type
\coq{A ~> B} to represent \emph{proper} morphisms of the
respectfulness relation of \coq{R} to \coq{R'}.
\begin{coqcode}
Record morph A {eA : setoid A} B {eB : setoid B} :=
  MkMorph { app :> A -> B; app_eq : forall x y, x =e y -> app x =e app y }.
Notation "A ~> B" := (@morph A _ B _).
\end{coqcode}
We rely on Coq's type class mechanism to fill in the necessary equivalence
relations. Coq's code extraction mechanism will erase any occurrence
of \coq{Prop} in the code, so objects of type \coq{A ~> B} will be extracted to
the OCaml equivalent to \coq{A -> B}. Note the implicit coercion
from \coq{A ~> B} to \coq{A -> B}.
On top of this, we define basic function composition and identity functions:
\begin{coqcode}
Notation "f \o g" = (comp f g).
Definition comp : (B ~> C) ~> (A ~> B) ~> A ~> C := ...
Definition id : A ~> A := ...
\end{coqcode}
Using custom equivalences and proper morphisms, we redefine the definitions of
container extensions and container equality. In particular, container
extensions require a proper morphism to check the validity of positions in
shapes, and container equality now uses equivalences of shapes and contained
elements: 
\begin{coqcode}
Inductive AppR F `{setoid X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{coqcode}
Note the use of \coq{=e} instead of Coq's standard equality.  For positions,
however, we chose to use Coq's propositional equality, since this leads
to simpler code. We explain why in Section~\ref{sec:algebras}, and the
mechanisation of initial algebras of container extensions.
%The definitions in
%Figure~\ref{fig:poly} are redefined to account for these changes in a
%straightforward way, i.e. we prove that validity predicates are indeed proper
%morphisms.

@@

\section{Formalising Recursion Schemes}
\label{sec:recursion-schemes}
Recursion schemes provide an abstract way to consume and generate data. We now
proceed onto describing how to formalise hylomorphisms in Coq. We first
formalise algebras for container extensions (Section~\ref{sec:algebras}), then
we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together
these notions to formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:rec-coalgebras}).

\subsection{Algebras and Catamorphisms for Containers}
\label{sec:algebras}

@@

\subsection{Coalgebras and Anamorphisms}
\label{sec:coalg}

@@

\subsection{Mechanising Hylomorphisms}
\label{sec:rec-coalgebras}
Recall that hylomorphisms capture the concept of \emph{divide-and-conquer}
algorithms where the input is first destructured (\emph{divide}) in smaller
parts by means of a coalgebra which are computed recursively and then composed
back together (\emph{conquer}) by means of an algebra.

@@

\noindent
It is important to highlight that, in our mechanisation, these proofs follow
\emph{exactly} the steps that one would do in a pen-and-paper proof.  We show
the series of rewrite steps for \coq{hylo_fusion_l} in
Figure~\ref{fig:proof-fusion-l}. The steps of this proof are: (1) apply the
uniqueness law of recursive hylomorphisms; (2) rewrite the goal using the
property that functors preserve composition; (3) rewrite the goal using the
condition of \coq{hylo_fusion_l}; (4) use the uniqueness law of hylomorphisms to
fold $a \circ F \; f \circ c$ into $f$.
\begin{figure}
\begin{coqcode}
(* E2 : f2 \o g1 =e g2 \o fmap f2 *)
(* ------------------------------------- *)
(* Goal : f2 \o hylo g1 h1 =e hylo g2 h1 *)
apply hylo_uniq.
       (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1        *)
rewrite fmap_comp.
(* rearranging by associativity *)
       (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
rewrite <- E2.
(* rearranging by associativity *)
       (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o fmap (hylo g1 h1)) \o h1)      *)
rewrite <- hylo_unroll.
       (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1                             *)
\end{coqcode}

@@

Similarly, the Coq value \coq{inr (true, false)} as a container
extension is as follows:
\begin{coqcode}
  {| shape := inr (tt,tt);
     cont := fun e : {p | valid (inr (tt, tt)) p} =>
        match proj1_sig e with
        | inl f => match f with end
        | inr (inl tt) => fun _ => true
        | inr (inr tt) => fun _ => false
        end
  |}
\end{coqcode}
Here, the position \coq{inl f} is impossible, as
\coq{f} would need to be a proof of \coq{False}.
Of the two remaining positions, the left one will return
\coq{true}, and the right position \coq{false}.
\begin{figure}
\begin{minipage}{.52\textwidth}
  \begin{minted}[linenos, fontsize=\small]{Coq}
Context `(C1 : Cont S1 P1)
        `(C2 : Cont S2 P2).

>>>>

\subsection{Types and Morphisms}
\label{sec:types-morphisms}
We address the different forms of equality by defining a class of
\emph{setoids}, types with an associated equivalence relation, and considering
only functions that respect the associated equivalences, or \emph{proper}
morphisms with respect to the function \emph{respectfulness} relation. We use
the type-class mechanism, instead of setoids in Coq's standard library, to help
Coq's code extraction mechanism remove any occurence of custom equivalence
relations in the extracted OCaml code. We use \coq{=e} to denote the equivalence
relation of a setoid. By default, we associate every Coq type with the standard
propositional equality, unless a different equivalence is specified (we allow
overlapping instances, and Coq's propositional equality takes the lowest
priority).
%\begin{coqcode}
%Reserved Notation "f =e g" (at level 70, no associativity).
%Class setoid A : Type :=
%  MkEquiv
%    { eqRel : A -> A -> Prop;
%      e_refl : forall x, eqRel x x;
%      e_sym : forall x y, eqRel x y -> eqRel y x;
%      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
%    }.
%Notation "f =e g" := (eqRel f g).
%\end{coqcode}
Given types \coq{A} and  \coq{B}, with their respective
equivalence relations \coq{eA : A -> A -> Prop} and
\coq{eB : B -> B -> Prop}, the we define the type
\coq{A ~> B} to represent \emph{proper} morphisms of the
respectfulness relation of \coq{R} to \coq{R'}.
\begin{coqcode}
Record morph A {eA : setoid A} B {eB : setoid B} :=
  MkMorph { app :> A -> B; app_eq : forall x y, x =e y -> app x =e app y }.
Notation "A ~> B" := (@morph A _ B _).
\end{coqcode}
We rely on Coq's type class mechanism to fill in the necessary equivalence
relations. Coq's code extraction mechanism will erase any occurrence
of \coq{Prop} in the code, so objects of type \coq{A ~> B} will be extracted to
the OCaml equivalent to \coq{A -> B}. Note the implicit coercion
from \coq{A ~> B} to \coq{A -> B}.
On top of this, we define basic function composition and identity functions:
\begin{coqcode}
Notation "f \o g" = (comp f g).
Definition comp : (B ~> C) ~> (A ~> B) ~> A ~> C := ...
Definition id : A ~> A := ...
\end{coqcode}
Using custom equivalences and proper morphisms, we redefine the definitions of
container extensions and container equality. In particular, container
extensions require a proper morphism to check the validity of positions in
shapes, and container equality now uses equivalences of shapes and contained
elements: 
\begin{coqcode}
Inductive AppR F `{setoid X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{coqcode}
Note the use of \coq{=e} instead of Coq's standard equality.  For positions,
however, we chose to use Coq's propositional equality, since this leads
to simpler code. We explain why in Section~\ref{sec:coq-algebras}, and the
mechanisation of initial algebras of container extensions.
%The definitions in
%Figure~\ref{fig:poly} are redefined to account for these changes in a
%straightforward way, i.e. we prove that validity predicates are indeed proper
%morphisms.

@@

\section{Formalising Recursion Schemes}
\label{sec:recursion-schemes}
Recursion schemes provide an abstract way to consume and generate data. We now
proceed onto describing how to formalise hylomorphisms in Coq. We first
formalise algebras for container extensions (Section~\ref{sec:coq-algebras}), then
we formalise coalgebras (Section~\ref{sec:coq-coalg}) and then we put together
these notions to formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:coq-rec-coalgebras}).

\subsection{Algebras and Catamorphisms for Containers}
\label{sec:coq-algebras}

@@

\subsection{Coalgebras and Anamorphisms}
\label{sec:coq-coalg}

@@

\subsection{Mechanising Hylomorphisms}
\label{sec:coq-rec-coalgebras}
Recall that hylomorphisms capture the concept of \emph{divide-and-conquer}
algorithms where the input is first destructured (\emph{divide}) in smaller
parts by means of a coalgebra which are computed recursively and then composed
back together (\emph{conquer}) by means of an algebra.

@@

\noindent
It is important to highlight that, in our mechanisation, these proofs follow
\emph{exactly} the steps that one would do in a pen-and-paper proof.  We show
the series of rewrite steps for \coq{hylo_fusion_l} in
Figure~\ref{fig:proof-fusion-l}. The steps of this proof are: (1) apply the
uniqueness law of recursive hylomorphisms; (2) rewrite the goal using the
property that functors preserve composition; (3) rewrite the goal using the
condition of \coq{hylo_fusion_l}; (4) use the uniqueness law of hylomorphisms to
fold $a \circ F \; f \circ c$ into $f$.
\begin{figure}[t!]
\begin{coqcode}
(* E2 : f2 \o g1 =e g2 \o fmap f2 *)
(* ------------------------------------- *)
(* Goal : f2 \o hylo g1 h1 =e hylo g2 h1 *)
apply hylo_uniq.
       (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1        *)
rewrite fmap_comp.
rewrite ... (* rearranging by associativity *)
       (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
rewrite <- E2.
rewrite ... (* rearranging by associativity *)
       (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o fmap (hylo g1 h1)) \o h1)      *)
rewrite <- hylo_unroll.
       (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1                             *)
\end{coqcode}

@@

Similarly, the Coq value \coq{inr (true, false)} as a container
extension is as follows:
\begin{coqcode}
  {| shape := inr (tt,tt);
     cont := fun e : {p | valid (inr (tt, tt)) p} =>
        match proj1_sig e with
        | inl f => match f with end
        | inr (inl tt) => fun _ => true
        | inr (inr tt) => fun _ => false
        end
  |}
\end{coqcode}
Here, the position \coq{inl f} is impossible, as
\coq{f} would need to be a proof of \coq{False}.
Of the two remaining positions, the left one will return
\coq{true}, and the right position \coq{false}.
\begin{figure}[t!]
\begin{minipage}{.52\textwidth}
  \begin{minted}[linenos, fontsize=\small]{Coq}
Context `(C1 : Cont S1 P1)
        `(C2 : Cont S2 P2).
